---
title: "Suicide Prevention and Analysis"
output: html_notebook
---


```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(factoextra)
library(MASS)
theme_set(theme_gray())
```


**Suicide Understanding and Prevention**

This notebook contains the analysis of a suicide dataset, which was the final
assignment of the Statistical data analysis course.

## 1. Exploratory data analysis

```{r include=FALSE}
# Load the dataset
load(file="data/suicide.RData")
```

I begin by inspecting whether the dataset contains any NA values that would need to be taken care of.
```{r}
data %>% is.na() %>% colSums()
```
Did all of the included countries contribute the same amount of data? Are there
any countries that reported their statistics far less often than other countries?

```{r}
years.count <- data %>% group_by(continent, country) %>%
  summarise(count = n_distinct(year)) %>%
  arrange(desc(count))

years.count %>%
  ggplot(aes(x = count, fill = continent)) +
  scale_x_continuous(n.breaks = 31) +
  geom_bar()
```


We can see that there are some countries that contributed only 3 years worth of
data or less.


### Suicide rate

Throughout the majority of this notebook, I will be working with suicide rates rather than with concrete numbers
of total suicides. The reason for this is that it allows us to perform better comparisons of suicide prevalence
between different countries and continents, as the populations of different areas differ, and the total number of
suicides in a population is correlated with the size of the population.

```{r}
country.summary <- data %>% group_by(year) %>%
  summarise(population = sum(population),
            suicides_no = sum(suicides_no))

cor(x = country.summary$population, y = country.summary$suicides_no)
```

If we wanted to be even more rigorous with our analysis, we could standardize our suicide rates by age distribution.


### Global trend inspection

The first thing that we can inspect is the progression of suicide rates worldwide during the time period 1985-2015.

```{r}

global.suicide.rate <- data %>% group_by(year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000
  ) 

global.suicide.rate %>%
  ggplot(aes(x = year, y = suicides_per_100k)) +
  geom_point(color="orange", size=2.4) +
  geom_line(color="orange", size=1.1) +
  labs(
    title = "Worldwide Suicide Rates",
    x = "Year", y = "Suicides per 100k people"
  ) +
  scale_x_continuous(breaks = seq(1985, 2015, 2))
  
```

This figure shows that there has been a great rise in suicides throughout the years 1989 and 1995.
However, with a few exceptions, the suicide rate seems to be on decline ever since then.

```{r}
differences <- data.frame(
  difference = tail(global.suicide.rate$suicides_per_100k, -1) - head(global.suicide.rate$suicides_per_100k, -1),
  year = tail(global.suicide.rate$year, -1))

differences <- differences %>%
  mutate(changeNotPositive = difference <= 0)

differences %>%
  ggplot(aes(x = year, y = difference, color = changeNotPositive)) +
  geom_segment(aes(x = year, xend = year, y = 0, yend = difference)) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(1986, 2015, 2)) +
  scale_y_continuous(breaks = seq(-1, 1.5, 0.5)) +
  theme(legend.position = "none") +
  labs(title = "Change in suicide rate in time",
       subtitle = "Shows the change in suicide rate from the previous year",
       x = "Year", y = "Difference")
```
The figure above shows us the differences in suicide rates between two subsequent years.

```{r fig.height=7}

continent_overall_plot <- data %>% group_by(continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = continent)) +
  geom_bar(stat = "identity") +
  labs(title = "Overall differences in suicide rates between continents",
       x = "Continent", y = "Suicides per 100k people") +
  theme(legend.position = "none")

continent_by_year_plot <- data %>% group_by(year, continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=continent)) +
  geom_line() +
  geom_point(size=0.8) +
  scale_x_continuous(breaks = seq(1985, 2015, 5)) +
  facet_wrap(~ continent, ncol = 1, scales = "free_y") +
  labs(title = "Suicide Rate By Continent In Time",
       x = "Year", y = "Suicides per 100k people") +
  theme(legend.position = "none")

grid.arrange(continent_overall_plot, continent_by_year_plot, ncol = 2)
```



```{r}
data %>%
  filter(continent == "Africa") %>%
  group_by(country) %>%
  summarise(reports = n_distinct(year)) %>%
  arrange(reports)
```

```{r fig.height = 6}

data_europe <- data %>% filter(continent == "Europe") %>%
  group_by(country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  arrange(desc(suicides_per_100k))

data_europe %>%
  ggplot(aes(x = reorder(country, suicides_per_100k), y = suicides_per_100k, fill = suicides_per_100k)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Overall Suicide Rates in European countries",
       x = "Country", y = "Suicides per 100k people")
  
```

```{r fig.height = 6}

data_europe %>%
  ggplot(aes(x = reorder(country, suicides_no), y = suicides_no, fill = suicides_no)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Number of suicides in European countries",
       x = "Country", y = "Number of suicides")
  
```

### Differences between sexes

```{r}
data %>% group_by(year, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=sex)) +
  geom_line(size=1.1) + 
  geom_point(size=2.4) +
  labs(title = "Global Differences in Suicide Rates Between Sexes",
       x = "Year", y = "Suicides per 100k people")
```

It is clear that the male part of the population is at a higher risk of suicide
than its female counterpart. Another somewhat interesting observation is that
the change in suicide rate of women in time is not as pronounced as is the
change of suicide rate of men.

```{r}
data %>% group_by(continent, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(breaks = seq(0, 40, 5)) +
  coord_flip() +
  theme(legend.position = "bottom") +
  labs(title = "Overall differences in suicide rates between the continents",
       subtitle = "along with differences between sexes within individual continents",
       x = "Continent", y = "Suicides per 100k people")
```
We can also inspect the differences between the sexes on within individual
continents. This again reveals the same pattern for all of the continents,
that men 

```{r fig.height=7, fig.width=7}
data %>% group_by(continent, sex, year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = sex)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ continent, ncol = 1, scales = "free_y", strip.position = "right") +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  theme(legend.position = "bottom")
```
The suicide rates of women are consistently lower than those of men across all
of the inspected continents.

### Age

```{r}
data %>% group_by(year, age) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point(size = 0.9) +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  scale_color_discrete() +
  labs(title = "Relationship between suicide rate and age",
       x = "Year", y = "Suicides per 100k people")
```

```{r fig.height=10, fig.width=8}
data %>% group_by(continent, year, age) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point(size = 0.9) +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  scale_color_discrete() +
  facet_wrap(~continent, ncol = 1) +
  labs(title = "Relationship between suicide rate and age within individual continents",
       x = "Year", y = "Suicides per 100k people")
```



### Correlation Between GDP and Suicide Rates

Using our dataset, we can also try to find out whether the gross domestic product
of a country affects its suicide rate. 

```{r}

data %>% group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    mean_gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep") %>%
  ggplot(aes(x = mean_gdp_per_capita, y = suicides_per_100k, color = continent)) +
  geom_point() +
  labs(title = "Relationship between suicide rate and GDP per capita",
       subtitle = "Both variables are represented as means from the time period 1985-2015",
       x = "GDP per capita [$]", y = "Suicides per 100k people")
  

```
This figure does not seem to reveal any relationship between these variables. We can also try to calculate the correlations
between these two variables over the same time interval for each of the countries individually. The correlations were computed using Spearman's rho, since the relationship between the two variables is not expected to be linear.
The correlations have been computed only for the countries where 10 or more years of data has been available.

```{r}

correlation.gdp.suicide <- data.frame()

for (c in levels(data$country)) {

  country.data <- data %>% filter(country == c) %>%
    group_by(country, year, gdp_per_capita) %>%
    summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep")
  
  if (nrow(country.data) < 10) {
    next
  }
  
  # corr <- cor(x = country.data$gdp_per_capita, y = country.data$suicides_per_100k, method = "spearman")
  corr <- cor.test(x = country.data$gdp_per_capita, y = country.data$suicides_per_100k, method = "spearman",
                   exact = FALSE)
  
  correlation.gdp.suicide <- rbind(
    correlation.gdp.suicide,
    list(
      country = c,
      correlation = corr$estimate,
      p.value = corr$p.value,
      significant = corr$p.value < 0.05
    )
  )
}


correlation.gdp.suicide <- arrange(correlation.gdp.suicide, correlation.gdp.suicide$correlation)

```

```{r}
head(correlation.gdp.suicide)
tail(correlation.gdp.suicide)
```

We can see that this provided us with a much more better insight into the data, and that there is truly a relationship between the suicide rate in a country and the gross domestic product of that country.
This relationship seems to be quite strong for a considerable amount of countries.
Some of the countries achieve a large positive correlation and some achieve a large negative correlation.
However, they are not distributed equally, as negative correlations are present more often than the positive ones.
All the correlations (with the exception of Cyprus) whose absolute value is larger than 0.4 are statistically significant with confidence level $\alpha = 0.05$.

```{r}
correlation.gdp.suicide %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = correlation)) +
  geom_histogram(binwidth = 0.1, center = 0.1) +
  scale_x_continuous(breaks = seq(-1, 1, 0.1)) +
  labs(title = "Histogram of statistically significant correlations of suicide rate and GDP")
```

Let's inspect this relationship further by plotting the suicide rate against GDP
for six of the countries that achieved the largest negative correlations.

```{r fig.width=8}

top.corr.countries <- head(correlation.gdp.suicide)$country
top.corr.correlations <- head(correlation.gdp.suicide)$correlation

my_labeller <- function(countries) {
  correlation <- map(
    countries,
    function(x) correlation.gdp.suicide[correlation.gdp.suicide$country == x, ]$correlation
  ) %>% unlist
  correlation <- round(correlation, digits = 2)
  result <- paste(as.character(countries), ", rho = ", correlation, sep = "")
}

data %>% filter(country %in% top.corr.countries) %>%
  group_by(country, year, gdp_per_capita) %>%
  summarise( 
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  mutate(country_order = factor(country, levels = top.corr.countries)) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = country_order)) +
  geom_point() +
  scale_colour_viridis_d(end = 0.8) +
  facet_wrap(~ country_order, scales = "free",
             labeller = as_labeller(my_labeller)) +
  labs(title = "Countries with the largest negative correlation of gross domestic product and suicide rate",
       subtitle = "Correlation is reported using Spearman's rho",
       x = "GDP [$]", y = "Suicides per 100k people") +
  theme(legend.position = "none")

```

We can see that plotting the data allows us to easily notice the relationship between these two variables.
It is also worth noting that all of these countries achieving the largest negative correlations are European.

We can do the same for countries that achieved the largest positive correlation.

```{r fig.width=8}

top.corr.countries <- head(arrange(correlation.gdp.suicide, desc(correlation.gdp.suicide$correlation)))
top.corr.countries <- top.corr.countries$country

my_labeller <- function(countries) {
  correlation <- map(
    countries,
    function(x) correlation.gdp.suicide[correlation.gdp.suicide$country == x, ]$correlation
  ) %>% unlist
  correlation <- round(correlation, digits = 2)
  result <- paste(as.character(countries), ", rho = ", correlation, sep = "")
}

data %>% filter(country %in% top.corr.countries) %>%
  group_by(country, year, gdp_per_capita) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  mutate(country_order = factor(country, levels = top.corr.countries)) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = country_order)) +
  geom_point() +
  scale_colour_viridis_d(begin = 0.8, end = 0) +
  facet_wrap(~ country_order, scales = "free",
             labeller = as_labeller(my_labeller)) +
  labs(title = "Countries with the largest correlation of gross domestic product and suicide rate",
       subtitle = "Correlation is reported using Spearman's rho",
       x = "GDP [$]", y = "Suicides per 100k people") +
  theme(legend.position = "none")
```

Once again, the interaction between suicide rate and GDP is easily recognizable, although in some cases
(e.g. Montenegro) it does not seem to be as well structured as it has been for the negative correlations.

Another thing I would like to investigate is whether there is a connection between the overall suicide rate
of a country and its dependence on the GDP of the country.


```{r}
data %>%
  filter(country %in% correlation.gdp.suicide$country) %>%
  group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep") %>%
  inner_join(correlation.gdp.suicide, by = "country") %>%
  # filter(p.value < 0.05) %>% 
  ggplot(aes(x = correlation, y = suicides_per_100k, color = continent, shape = significant)) +
  geom_point() +
  labs(title = "Relationship between suicide rate and its correlation with GDP",
       subtitle = "Correlation is reported using Spearman's Rho",
       x = "Correlation", y = "Suicides per 100k people")
```
The plot includes the correlations even if they were not statistically significant.
It seems that the countries, where the suicide rate is highly correlated with the gross domestic
product, tend to have higher suicide rates than the countries where the correlation with GDP
is not very strong. This can be observed on both ends of the correlation's value range, however it is
far more pronounced for countries that achieve a negative correlation.

We can also observe that the suicide rates of countries that achieve a large negative correlation tend
to be larger than of countries that achieve a large positive correlation.

Once again, it is worth noting that the majority of countries that achieve a negative correlation is European.
Let's see what correlations did individual European countries achieve.

```{r fig.height=4, fig.width=8}

countries.continent <- distinct(data[c("country", "continent")])

correlation.gdp.suicide %>%
  inner_join(countries.continent, by = "country") %>%
  filter(continent == "Europe") %>%
  ggplot(aes(x = reorder(country, correlation), y = correlation, fill = correlation)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.1)) +
  labs(title = "Correlations of European countries",
       x = "Country", y = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none")

```

We can see that most of the European countries truly achieve large negative correlations of
suicide rates and GDP. However, there are a few exceptions, where these two variables
are correlated positively.


```{r}
data_japan <- data %>% filter(country == "Japan")

data_japan %>% group_by(country, year) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = country)) +
  geom_line() +
  geom_point() +
  theme(legend.position = "none")

data_japan %>% group_by(country, year, age) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  labs(title = "Differences between age groups in Japan in time",
       x = "Year", y = "Suicides per 100k people")

```


### Clustering

```{r}
clustering.data <- data %>%
  filter(country %in% correlation.gdp.suicide$country) %>%
  group_by(country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita)) %>%
  inner_join(correlation.gdp.suicide, by = "country") %>%
  mutate(suicides_per_100k = scale(suicides_per_100k),
         correlation = scale(correlation)) %>%
  dplyr::select("country", "correlation", "suicides_per_100k")

clustering.data.numeric <- clustering.data %>% dplyr::select(correlation, suicides_per_100k)

clustering <- kmeans(clustering.data.numeric, 6)

fviz_cluster(clustering, data = clustering.data.numeric, labelsize = 0) +
  labs(title = "Clustering countries based on their dependence on GDP",
       subtitle = "Both features have been normalized",
       x = "GDP and Suicide rate correlation")


```

```{r}
clustering.data %>% cbind(cluster = clustering$cluster) %>%
  filter(cluster == 5) %>%
  pull(country)
```

```{r}
clustering
```


### Dimension Reduction

The $\texttt{generation}$ feature is redundant, since we can assign a generation to a group of population by combining the information
from the features $\texttt{year}$ and $\texttt{age}$.

Furthermore, the features $\texttt{gdp_for_year}$ and $\texttt{gdp_per_capita}$ are mutually redundant. That is because we can always
use one of these features in combination with the sum of total population in a country in a certain year to get the value of the other
feature, e.g. $\texttt{gdp_per_capita} = \frac{\texttt{gdp_for_year}}{\texttt{total_population}}$.

```{r}
data %>% group_by(country, year, gdp_for_year, gdp_per_capita) %>%
  summarise(population = sum(population)) %>%
  mutate(x = as.integer(round(gdp_for_year / population)),
         y = abs(x - gdp_per_capita)) %>%
  pull(y) %>%
  mean
```


### Conclusion

We have uncovered many things during the exploratory data analysis. First of all, we have seen that
the world experienced a sudden rise in suicide rates during 1989-1995, but these have been on
decline since then. We also inspected the difference in suicide rates between the individual continents.
This revealed that Europe has the highest suicide rate, followed by Asia. Inspection of the suicide rates
in European countries shown us that Lithuania has the highest suicide rate, however the number of suicides
has been the largest in Russia.

We have also seen that women 

## Hypotheses Testing

Based on the previous analysis I would like to test several hypotheses.
The confidence level is set to be $\alpha = 0.05$.

### Suicide rates of men and women

The first hypothesis that I will test is that that men tend to commit
suicide more often than women. Therefore, I will test if the suicide rate of men
was higher than that of women during the time interval 1985-2015.

Let $\mu_m$ and $\mu_w$ be the means of suicide rates of men and women, respectively.

The null hypothesis is then
$$ \mu_m \leq \mu_w $$ 
and the alternative hypothesis is
$$ \mu_m > \mu_w $$
where $\mu_m$ and $\mu_w$ are the means of suicide rates of men and women during
the time period 1985-2015, respectively. The confidence level is set to be $\alpha = 0.05$.

In order to test this hypothesis, I will make use of Welch's Two Sample t-test,

In order to test this hypothesis, I will make use of the Welch's Two Sample t-test,
which is a modification of Student's t-test supporting unequal variances of the samples.


```{r}

sex.hypothesis.data <- data %>% group_by(year, sex) %>%
    summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
    )

rates.male <- sex.hypothesis.data %>% filter(sex == "Male") %>% pull(suicides_per_100k)
rates.female <- sex.hypothesis.data %>% filter(sex == "Female") %>% pull(suicides_per_100k)

t.test(
  x = rates.male,
  y = rates.female,
  alternative = "greater"
)

```

We can see that the resulting p-value of the Welch's Two Sample t-test
is smaller than our confidence level, and because of that we can reject the null hypothesis
in favor of the alternative hypothesis. We have confirmed our initial suspicion,
that is that men have higher suicide rates than women.


### Difference in suicide rates between age groups

Another interesting insight gained from the exploratory analysis was that there
is an apparent difference in suicide rates between individual age groups.

Usually, we could use ANOVA to test if at least one of the groups differs significantly
from the rest of the groups. Unfortunately, our data violates certain assumptions of this method,
namely the normality assumption, which requires every group to be normally distributed.

We can easily show the violation of this assumption by inspecting one of the age groups,
for example the group containing people older than 75 years. We can check whether it is
normally distributed by taking a look at the histogram of differences of suicide rates
throughout the years and their mean.

```{r}

age.old.rates <- data %>%
  filter(age == "75+") %>%
  group_by(year) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000)

(data.frame(residual = age.old.rates$suicides_per_100k - mean(age.old.rates$suicides_per_100k))) %>%
  ggplot(aes(x = residual)) +
  scale_fill_viridis_c() +
  geom_histogram(binwidth = 2,
                 center = 1,
                 fill = viridis::viridis(1, begin = 0.7),
                 color = viridis::viridis(1, begin = 0.8)) +
  labs(title = "Histogram of differences of yearly suicide rates and their mean",
       subtitle = "for people that are 75+ years old",
       x = "Residual", y = "Count")

```

This histogram indicates that the distribution of suicide rates in fact does not originate
from the normal distribution.

Due to this assumption being violated, we are forced to use a different method,
the Kruskal-Wallis test, which can be used even if the normality assumption does not hold.
This method will allow us to test a similar hypothesis.



```{r}

age.hypothesis.data <- data %>%
  group_by(year, age) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
    )

kruskal.test(suicides_per_100k ~ age, data = age.hypothesis.data)

```


### Higher suicide rates in countries with negative correlation

During the analysis we have seen that countries that achieve large negative correlation have
a larger suicide rate than countries that achieve a large positive correlation. It might be
interesting to test this hypothesis.

Let us create two subsets of the countries based on their correlations. The first group will contain
countries that achieved correlations smaller than $-0.6$, and the other group will contain countries
whose suicide rate and GDP correlation was larger than $0.6$.

Let $\mu_l$ and $mu_h$ be the mean suicide rates of countries of the first and second group, respectively.

The null hypothesis is then
$$ \mu_l \leq \mu_h $$
and the alternative hypothesis is
$$ \mu_l > \mu_h $$

Once again, we will make use of the Welch's Two Sample t-test.


```{r}
signif.corr <- data %>%
  filter(country %in% correlation.gdp.suicide$country) %>%
  group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep") %>%
  inner_join(correlation.gdp.suicide, by = "country") %>%
  filter(p.value < 0.05)

t.test(x = signif.corr %>% filter(correlation < -0.6) %>% pull(suicides_per_100k),
       y = signif.corr %>% filter(correlation > 0.6) %>% pull(suicides_per_100k),
       alternative = "greater")
  
```

The resulting p-value is less than our predefined confidence level, therefore we reject the null hypothesis
in favor of the alternative hypothesis.

The other hypothesis I would like to test is that countries with large absolute values of the suicide rate
and GDP correlation have higher suicide rates than countries that have a small absolute correlation.
However, I did not want to base this hypothesis on the data contained in this dataset, since the small correlations did not turn out to be statistically significant.

