---
title: "R Notebook"
output: html_notebook
---


```{r setup, include=FALSE}
library(tidyverse)
library(MASS)
theme_set(theme_gray())
```


**Suicide Understanding and Prevention**

This notebook contains the analysis of a suicide dataset, which was the final
assignment of the Statistical data analysis course.

## 1. Exploratory data analysis

```{r include=FALSE}
# Load the dataset
load(file="data/suicide.RData")
```

I begin by inspecting whether the dataset contains any NA values that would be
needed to be taken care of.
```{r}
data %>% is.na() %>% colSums()
```
Did all of the included countries contribute the same amount of data? Are there
any countries that reported their statistics far less often than other countries?

```{r}
data %>% group_by(country) %>%
  summarise(years_count = n_distinct(year), .groups="keep") %>%
  arrange(desc(years_count))
```

We can see that there are some countries that contributed only 3 years worth of
data or less.

```{r}
data %>% group_by(year) %>% summarise(population = sum(population), suicides_no = sum(suicides_no), per_100k = (suicides_no / population) * 100000)
```

### Global trend inspection

We can inspect the global trend of suicide rates.

```{r}
global_rate <- data %>% group_by(year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  )

gg <- ggplot(global_rate, aes(x = year, y = suicides_per_100k)) +
  geom_point(color="orange", size=2.4) +
  geom_line(color="orange", size=1.1) +
  labs(
    title = "Global Suicide Rates",
    x = "Year", y = "Suicides per 100k people"
  ) +
  scale_x_continuous(breaks = seq(1985, 2015, 2))
  

plot(gg)
```
```{r}
data %>% group_by(year, continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=continent)) +
  geom_line() +
  geom_point(size=0.8) +
  facet_wrap(~ continent)
```
```{r}
data %>%
  filter(continent == "Africa") %>%
  group_by(country) %>%
  summarise(reports = n_distinct(year)) %>%
  arrange(reports)
```
```{r}
data %>% filter(country == "Cabo Verde") %>%
  summarise(min_year = min(year), max_year = max(year))
```


```{r}
# Global rates conditioned on gender

gg <- data %>% group_by(year, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=sex)) +
  geom_line(size=1.1) + 
  geom_point(size=2.4) +
  labs(title = "Global Differences in Suicide Rates Between Sexes",
       x = "Year", y = "Suicides per 100k people")
  
plot(gg)

```

It is clear that the male part of the population is at a higher risk of suicide
than its female counterpart. Another somewhat interesting observation is that
the change in suicide rates of women in time is not as pronounced as is the
change of suicide rate of men.

```{r}
data %>% group_by(continent, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(breaks = seq(0, 40, 5)) +
  coord_flip() +
  labs(title = "Overall Differences in Suicide Rates between the Continents",
       subtitle = "Along with Differences Between Sexes within Individual Continents",
       x = "Continent", y = "Suicides per 100k people")
```



```{r}
data %>% filter(year == 1990) %>%
  group_by(country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    gdp_per_capita = gdp_per_capita,
    continent = continent,
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = continent)) +
  scale_x_log10() +
  geom_point()
```

```{r}
gg <- data %>% group_by(year, age) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  labs(title = "", x = "Year", y = "Suicides per 100k people")

plot(gg)
  
```

```{r fig.height = 6}
data %>% filter(continent == "Europe") %>%
  group_by(country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  arrange(desc(suicides_per_100k)) %>%
  ggplot(aes(x = reorder(country, suicides_per_100k), y = suicides_per_100k, fill = suicides_per_100k)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "blue", high = "red") + 
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Overall Suicide Rates in European countries",
       x = "Country", y = "Suicides per 100k people")
  
```

```{r}

pca <- prcomp(
  data %>%
    group_by(country) %>%
    summarise(
      mean_gdp = mean(gdp_per_capita),
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
    ) %>% select(suicides_per_100k, mean_gdp))

```
```{r}

data.for.lda <- data %>%
  group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000)

data.lda <- lda(suicides_per_100k ~ continent, data = data.for.lda)
```


```{r}
data_pca <- data %>%
  group_by(country) %>%
  summarise(
    mean_gdp = mean(gdp_per_capita),
    population = sum(population),
    suicides_no = sum(suicides_no),
    mean_gdp_year = mean(gdp_for_year),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep")

pca <- prcomp(data_pca[c("suicides_per_100k", "mean_gdp")], center = TRUE, scale. = TRUE)
show(pca)

biplot(pca)
```

