---
title: "R Notebook"
output: html_notebook
---


```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(MASS)
theme_set(theme_gray())
```


**Suicide Understanding and Prevention**

This notebook contains the analysis of a suicide dataset, which was the final
assignment of the Statistical data analysis course.

## 1. Exploratory data analysis

```{r include=FALSE}
# Load the dataset
load(file="data/suicide.RData")
```

I begin by inspecting whether the dataset contains any NA values that would be
needed to be taken care of.
```{r}
data %>% is.na() %>% colSums()
```
Did all of the included countries contribute the same amount of data? Are there
any countries that reported their statistics far less often than other countries?

```{r}
data %>% group_by(country) %>%
  summarise(years_count = n_distinct(year), .groups="keep") %>%
  arrange(desc(years_count))
```

We can see that there are some countries that contributed only 3 years worth of
data or less.

```{r}
data %>% group_by(year) %>% summarise(population = sum(population), suicides_no = sum(suicides_no), per_100k = (suicides_no / population) * 100000)
```

### Global trend inspection

We can inspect the global trend of suicide rates.

```{r}
data %>% group_by(year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) +
  geom_point(color="orange", size=2.4) +
  geom_line(color="orange", size=1.1) +
  labs(
    title = "Global Suicide Rates",
    x = "Year", y = "Suicides per 100k people"
  ) +
  scale_x_continuous(breaks = seq(1985, 2015, 2))
  
```
```{r}
 data %>% group_by(year, continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=continent)) +
  geom_line() +
  geom_point(size=0.8) +
  facet_wrap(~ continent)
```

```{r fig.height=7}

continent_overall_plot <- data %>% group_by(continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = continent)) +
  geom_bar(stat = "identity") +
  labs(title = "Overall differences in suicide rates between continents",
       x = "Continent", y = "Suicides per 100k people") +
  theme(legend.position = "none")

continent_by_year_plot <- data %>% group_by(year, continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=continent)) +
  geom_line() +
  geom_point(size=0.8) +
  scale_x_continuous(breaks = seq(1985, 2015, 5)) +
  facet_wrap(~ continent, ncol = 1, scales = "free_y") +
  labs(title = "Suicide Rate By Continent In Time",
       x = "Year", y = "Suicides per 100k people") +
  theme(legend.position = "none")

grid.arrange(continent_overall_plot, continent_by_year_plot, ncol = 2)
```
```{r}
data %>%
  filter(continent == "Africa") %>%
  group_by(country) %>%
  summarise(reports = n_distinct(year), .groups = "keep") %>%
  arrange(reports)
```


### Differences between sexes

```{r}
# Global rates conditioned on gender

gg <- data %>% group_by(year, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=sex)) +
  geom_line(size=1.1) + 
  geom_point(size=2.4) +
  labs(title = "Global Differences in Suicide Rates Between Sexes",
       x = "Year", y = "Suicides per 100k people")
  
plot(gg)

```

It is clear that the male part of the population is at a higher risk of suicide
than its female counterpart. Another somewhat interesting observation is that
the change in suicide rate of women in time is not as pronounced as is the
change of suicide rate of men.

```{r}
data %>% group_by(continent, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(breaks = seq(0, 40, 5)) +
  coord_flip() +
  theme(legend.position = "bottom") +
  labs(title = "Overall Differences in Suicide Rates between the Continents",
       subtitle = "Along with Differences Between Sexes within Individual Continents",
       x = "Continent", y = "Suicides per 100k people")
```
We can also inspect the differences between the sexes on within individual
continents. This again reveals the same pattern for all of the continents,
that men 

```{r fig.height=7, fig.width=7}
data %>% group_by(continent, sex, year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = sex)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ continent, ncol = 1, scales = "free_y", strip.position = "right") +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  theme(legend.position = "bottom")
```
The suicide rates of women are consistently lower than those of men across all
of the inspected continents.


```{r}
data %>% group_by(continent, country, year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    gdp_per_capita = mean(gdp_per_capita),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = continent)) +
  scale_x_log10() +
  geom_point()
```
```{r}

data.lda <- data %>% group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    mean_gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep")

cor(data.lda$suicides_per_100k, data.lda$mean_gdp_per_capita, method = "spearman")

data.lda %>%
  ggplot(aes(x = mean_gdp_per_capita, y = suicides_per_100k, color = continent)) +
  geom_point()

data.glm <- glm(suicides_per_100k ~ poly(mean_gdp_per_capita, 2), data = data.lda)
summary(data.glm)

```

### Correlation Between GDP and Suicide Rates
```{r}

correlation.gdp.suicide <- data.frame()

for (c in levels(data$country)) {

  country.data <- data %>% filter(country == c) %>%
    group_by(country, year, gdp_per_capita) %>%
    summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep")
  
  if (nrow(country.data) < 10) {
    next
  }
  
  corr <- cor(x = country.data$gdp_per_capita, y = country.data$suicides_per_100k, method = "spearman")
  
  correlation.gdp.suicide <- rbind(
    correlation.gdp.suicide,
    list(
      country = c,
      correlation = corr
    )
  )
}


correlation.gdp.suicide <- arrange(correlation.gdp.suicide, correlation.gdp.suicide$correlation)

```

```{r}

correlation.gdp.suicide %>%
  filter(country %in% (data %>% filter(continent == "Africa") %>% distinct(country))$country) %>%
  ggplot(aes(x = reorder(country, correlation), y = correlation, fill = correlation)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_viridis_c()

```


```{r fig.width=8}

top.corr.countries <- head(correlation.gdp.suicide)$country
top.corr.correlations <- head(correlation.gdp.suicide)$correlation

my_labeller <- function(countries) {
  correlation <- map(
    countries,
    function(x) correlation.gdp.suicide[correlation.gdp.suicide$country == x, ]$correlation
  ) %>% unlist
  correlation <- round(correlation, digits = 2)
  result <- paste(as.character(countries), ", rho = ", correlation, sep = "")
}

data %>% filter(country %in% top.corr.countries) %>%
  group_by(country, year, gdp_per_capita) %>%
  summarise( 
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  mutate(country_order = factor(country, levels = top.corr.countries)) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = country_order)) +
  geom_point() +
  scale_colour_viridis_d(end = 0.9) +
  facet_wrap(~ country_order, scales = "free",
             labeller = as_labeller(my_labeller)) +
  labs(title = "Countries with the highest anticorrelation of gross domestic product and suicide rate",
       subtitle = "Correlation is reported using Spearman's rho",
       x = "GDP [$]", y = "Suicides per 100k people") +
  theme(legend.position = "none")

```

The correlations were computed using Spearman's rho, since the relationship between
the two variables does not appear to be linear.

```{r fig.width=8}

top.corr.countries <- head(arrange(correlation.gdp.suicide, desc(correlation.gdp.suicide$correlation)))
top.corr.countries <- top.corr.countries$country

my_labeller <- function(countries) {
  correlation <- map(
    countries,
    function(x) correlation.gdp.suicide[correlation.gdp.suicide$country == x, ]$correlation
  ) %>% unlist
  correlation <- round(correlation, digits = 2)
  result <- paste(as.character(countries), ", rho = ", correlation, sep = "")
}

data %>% filter(country %in% top.corr.countries) %>%
  group_by(country, year, gdp_per_capita) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  mutate(country_order = factor(country, levels = top.corr.countries)) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = country_order)) +
  geom_point() +
  scale_colour_viridis_d(begin = 0.9, end = 0) +
  facet_wrap(~ country_order, scales = "free",
             labeller = as_labeller(my_labeller)) +
  labs(title = "Countries with the highest correlation of gross domestic product and suicide rate",
       subtitle = "Correlation is reported using Spearman's rho",
       x = "GDP [$]", y = "Suicides per 100k people") +
  theme(legend.position = "none")
```

```{r}

data %>%
  filter(country %in% correlation.gdp.suicide$country) %>%
  group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep") %>%
  mutate(gdp.corr = correlation.gdp.suicide[correlation.gdp.suicide$country == country, ]$correlation) %>%
  ggplot(aes(x = gdp.corr, y = suicides_per_100k, color = country)) +
  scale_color_viridis_d() +
  geom_point() +
  theme(legend.position = "none") +
  labs(title = "Relationship between suicide rate and its correlation with GDP",
       subtitle = "Correlation is reported using Spearman's Rho",
       x = "correlation", y = "Suicides per 100k people")

```
It seems that the countries, where the suicide rate is highly correlated with the gross domestic
product, tend to have higher suicide rates than the countries where the correlation with GDP
is not very strong. This can be observed on both ends of the correlation domain, however it is


```{r}
gg <- data %>% group_by(year, age) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point(size = 0.9) +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  labs(title = "", x = "Year", y = "Suicides per 100k people")

plot(gg)
  
```
According to this graph, the suicide risk is 

```{r fig.height = 6}

data_europe <- data %>% filter(continent == "Europe") %>%
  group_by(country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  arrange(desc(suicides_per_100k))

data_europe %>%
  ggplot(aes(x = reorder(country, suicides_per_100k), y = suicides_per_100k, fill = suicides_per_100k)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Overall Suicide Rates in European countries",
       x = "Country", y = "Suicides per 100k people")
  
```

```{r}

european_countries <- data_europe %>% head(5) %>% dplyr::select(country)

data %>% filter(country %in% european_countries$country) %>%
  group_by(country, year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = country)) +
  geom_line()

```
```{r}

ru_be_data <- data %>% filter(country %in% c("Russian Federation", "Belarus")) %>%
  group_by(country, year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  )


```


```{r}

russia <- data %>% filter(country == "Russian Federation")

russia %>% group_by(year, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = sex)) +
  geom_line() +
  geom_point() +
  labs(title = "Suicide Rate in Russia In Time",
       x = "Year", y = "Suicides per 100k people")

```
Correlation between suicide rate and alcohol consumption.

```{r}

pca <- prcomp(
  data %>%
    group_by(country) %>%
    summarise(
      mean_gdp = mean(gdp_per_capita),
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
    ) %>% select(suicides_per_100k, mean_gdp))

```
```{r}
data_japan <- data %>% filter(country == "Japan")

data_japan %>% group_by(country, year) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = country)) +
  geom_line() +
  geom_point() +
  theme(legend.position = "none")

data_japan %>% group_by(country, year, age) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  labs(title = "Differences between age groups in Japan in time",
       x = "Year", y = "Suicides per 100k people")

```


```{r}
data_pca <- data %>%
  group_by(country) %>%
  summarise(
    mean_gdp = mean(gdp_per_capita),
    population = sum(population),
    suicides_no = sum(suicides_no),
    mean_gdp_year = mean(gdp_for_year),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep")

pca <- prcomp(data_pca[c("suicides_per_100k", "mean_gdp", "population")], center = TRUE, scale. = TRUE)
show(pca)

biplot(pca)
```

```{r}



```


