---
title: "R Notebook"
output: html_notebook
---


```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(MASS)
theme_set(theme_gray())
```


**Suicide Understanding and Prevention**

This notebook contains the analysis of a suicide dataset, which was the final
assignment of the Statistical data analysis course.

## 1. Exploratory data analysis

```{r include=FALSE}
# Load the dataset
load(file="data/suicide.RData")
```

I begin by inspecting whether the dataset contains any NA values that would be
needed to be taken care of.
```{r}
data %>% is.na() %>% colSums()
```
Did all of the included countries contribute the same amount of data? Are there
any countries that reported their statistics far less often than other countries?

```{r}
data %>% group_by(country) %>%
  summarise(years_count = n_distinct(year), .groups="keep") %>%
  arrange(desc(years_count))
```

We can see that there are some countries that contributed only 3 years worth of
data or less.

```{r}
data %>% group_by(year) %>% summarise(population = sum(population), suicides_no = sum(suicides_no), per_100k = (suicides_no / population) * 100000)
```

### Global trend inspection

We can inspect the global trend of suicide rates.

```{r}
data %>% group_by(year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k)) +
  geom_point(color="orange", size=2.4) +
  geom_line(color="orange", size=1.1) +
  labs(
    title = "Global Suicide Rates",
    x = "Year", y = "Suicides per 100k people"
  ) +
  scale_x_continuous(breaks = seq(1985, 2015, 2))
  
```
```{r fig.height=7}

continent_overall_plot <- data %>% group_by(continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = continent)) +
  geom_bar(stat = "identity") +
  labs(title = "Overall differences in suicide rates between continents",
       x = "Continent", y = "Suicides per 100k people") +
  theme(legend.position = "none")

continent_by_year_plot <- data %>% group_by(year, continent) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=continent)) +
  geom_line() +
  geom_point(size=0.8) +
  scale_x_continuous(breaks = seq(1985, 2015, 5)) +
  facet_wrap(~ continent, ncol = 1, scales = "free_y") +
  labs(title = "Suicide Rate By Continent In Time",
       x = "Year", y = "Suicides per 100k people") +
  theme(legend.position = "none")

grid.arrange(continent_overall_plot, continent_by_year_plot, ncol = 2)
```
```{r}
data %>%
  filter(continent == "Africa") %>%
  group_by(country) %>%
  summarise(reports = n_distinct(year), .groups = "keep") %>%
  arrange(reports)
```

```{r fig.height = 6}

data_europe <- data %>% filter(continent == "Europe") %>%
  group_by(country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  arrange(desc(suicides_per_100k))

data_europe %>%
  ggplot(aes(x = reorder(country, suicides_per_100k), y = suicides_per_100k, fill = suicides_per_100k)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Overall Suicide Rates in European countries",
       x = "Country", y = "Suicides per 100k people")
  
```

```{r fig.height = 6}

data_europe %>%
  ggplot(aes(x = reorder(country, suicides_no), y = suicides_no, fill = suicides_no)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Overall Suicide Rates in European countries",
       x = "Country", y = "Suicides per 100k people")
  
```

### Differences between sexes

```{r}
# Global rates conditioned on gender

gg <- data %>% group_by(year, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color=sex)) +
  geom_line(size=1.1) + 
  geom_point(size=2.4) +
  labs(title = "Global Differences in Suicide Rates Between Sexes",
       x = "Year", y = "Suicides per 100k people")
  
plot(gg)

```

It is clear that the male part of the population is at a higher risk of suicide
than its female counterpart. Another somewhat interesting observation is that
the change in suicide rate of women in time is not as pronounced as is the
change of suicide rate of men.

```{r}
data %>% group_by(continent, sex) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = continent, y = suicides_per_100k, fill = sex)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(breaks = seq(0, 40, 5)) +
  coord_flip() +
  theme(legend.position = "bottom") +
  labs(title = "Overall differences in suicide rates between the continents",
       subtitle = "along with differences between sexes within individual continents",
       x = "Continent", y = "Suicides per 100k people")
```
We can also inspect the differences between the sexes on within individual
continents. This again reveals the same pattern for all of the continents,
that men 

```{r fig.height=7, fig.width=7}
data %>% group_by(continent, sex, year) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = sex)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ continent, ncol = 1, scales = "free_y", strip.position = "right") +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  theme(legend.position = "bottom")
```
The suicide rates of women are consistently lower than those of men across all
of the inspected continents.

### Age

```{r}
data %>% group_by(year, age) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point(size = 0.9) +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  labs(title = "Relationship between suicide rate and age",
       x = "Year", y = "Suicides per 100k people")
```

```{r fig.height=10, fig.width=8}
data %>% group_by(continent, year, age) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point(size = 0.9) +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  facet_wrap(~continent, ncol = 1) +
  labs(title = "Relationship between suicide rate and age within individual continents",
       x = "Year", y = "Suicides per 100k people")
```



### Correlation Between GDP and Suicide Rates

Using our dataset, we can also try to find out whether the gross domestic product
of a country affects its suicide rate. 

```{r}

data %>% group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    mean_gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep") %>%
  ggplot(aes(x = mean_gdp_per_capita, y = suicides_per_100k, color = continent)) +
  geom_point() +
  labs(title = "Relationship between suicide rate and GDP per capita",
       subtitle = "Both variables are represented as means from the time interval 1985-2015",
       x = "GDP per capita [$]", y = "Suicides per 100k people")
  

```
This figure does not seem to reveal any .
We can also try to calculate the correlations between these two variables over the same time interval
for each of the countries individually.

The correlations were computed using Spearman's rho, since the relationship between
the two variables does not appear to be linear.

```{r}

correlation.gdp.suicide <- data.frame()

for (c in levels(data$country)) {

  country.data <- data %>% filter(country == c) %>%
    group_by(country, year, gdp_per_capita) %>%
    summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep")
  
  if (nrow(country.data) < 10) {
    next
  }
  
  corr <- cor(x = country.data$gdp_per_capita, y = country.data$suicides_per_100k, method = "spearman")
  
  correlation.gdp.suicide <- rbind(
    correlation.gdp.suicide,
    list(
      country = c,
      correlation = corr
    )
  )
}


correlation.gdp.suicide <- arrange(correlation.gdp.suicide, correlation.gdp.suicide$correlation)

```

```{r}
correlation.gdp.suicide %>%
  ggplot(aes(x = correlation)) +
  geom_density(color = "orange", fill = "orange", alpha = 0.2)
```


```{r fig.width=8}

top.corr.countries <- head(correlation.gdp.suicide)$country
top.corr.correlations <- head(correlation.gdp.suicide)$correlation

my_labeller <- function(countries) {
  correlation <- map(
    countries,
    function(x) correlation.gdp.suicide[correlation.gdp.suicide$country == x, ]$correlation
  ) %>% unlist
  correlation <- round(correlation, digits = 2)
  result <- paste(as.character(countries), ", rho = ", correlation, sep = "")
}

data %>% filter(country %in% top.corr.countries) %>%
  group_by(country, year, gdp_per_capita) %>%
  summarise( 
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  mutate(country_order = factor(country, levels = top.corr.countries)) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = country_order)) +
  geom_point() +
  scale_colour_viridis_d(end = 0.9) +
  facet_wrap(~ country_order, scales = "free",
             labeller = as_labeller(my_labeller)) +
  labs(title = "Countries with the highest anticorrelation of gross domestic product and suicide rate",
       subtitle = "Correlation is reported using Spearman's rho",
       x = "GDP [$]", y = "Suicides per 100k people") +
  theme(legend.position = "none")

```

It is also worth noting that all of these
countries are European.

```{r fig.width=8}

top.corr.countries <- head(arrange(correlation.gdp.suicide, desc(correlation.gdp.suicide$correlation)))
top.corr.countries <- top.corr.countries$country

my_labeller <- function(countries) {
  correlation <- map(
    countries,
    function(x) correlation.gdp.suicide[correlation.gdp.suicide$country == x, ]$correlation
  ) %>% unlist
  correlation <- round(correlation, digits = 2)
  result <- paste(as.character(countries), ", rho = ", correlation, sep = "")
}

data %>% filter(country %in% top.corr.countries) %>%
  group_by(country, year, gdp_per_capita) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
    .groups = "keep") %>%
  mutate(country_order = factor(country, levels = top.corr.countries)) %>%
  ggplot(aes(x = gdp_per_capita, y = suicides_per_100k, color = country_order)) +
  geom_point() +
  scale_colour_viridis_d(begin = 0.9, end = 0) +
  facet_wrap(~ country_order, scales = "free",
             labeller = as_labeller(my_labeller)) +
  labs(title = "Countries with the highest correlation of gross domestic product and suicide rate",
       subtitle = "Correlation is reported using Spearman's rho",
       x = "GDP [$]", y = "Suicides per 100k people") +
  theme(legend.position = "none")
```


```{r}

data %>%
  filter(country %in% correlation.gdp.suicide$country) %>%
  group_by(continent, country) %>%
  summarise(
    population = sum(population),
    suicides_no = sum(suicides_no),
    suicides_per_100k = (suicides_no / population) * 100000,
    gdp_per_capita = mean(gdp_per_capita),
    .groups = "keep") %>%
  mutate(gdp.corr = correlation.gdp.suicide[correlation.gdp.suicide$country == country, ]$correlation) %>%
  ggplot(aes(x = gdp.corr, y = suicides_per_100k, color = continent)) +
  geom_point() +
  labs(title = "Relationship between suicide rate and its correlation with GDP",
       subtitle = "Correlation is reported using Spearman's Rho",
       x = "Correlation", y = "Suicides per 100k people")

```
It seems that the countries, where the suicide rate is highly correlated with the gross domestic
product, tend to have higher suicide rates than the countries where the correlation with GDP
is not very strong. This can be observed on both ends of the correlation's value range, however it is
far more pronounced for countries that anticorrelate.
It is also apparent that the majority of countries that achieve a high correlation is European.

```{r fig.height=4, fig.width=8}

countries.continent <- distinct(data[c("country", "continent")])

correlation.gdp.suicide %>%
  inner_join(countries.continent, by = "country") %>%
  filter(continent == "Europe") %>%
  ggplot(aes(x = reorder(country, correlation), y = correlation, fill = correlation)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c() +
  scale_y_continuous(limits = c(-1, 1)) +
  labs(title = "Correlations of European countries",
       x = "Country", y = "Correlation") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "none")

# correlation.gdp.suicide %>%
#   inner_join(countries.continent, by = "country") %>%
#   filter(continent == "Europe") %>%
#   ggplot(aes(y = reorder(country, correlation), x = correlation, fill = continent)) +
#   geom_bar(stat = "identity") +
#   labs(title = "Correlations between suicide rate and gross domestic product")
#   # theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

correlation.gdp.suicide %>%
  inner_join(countries.continent, by = "country") %>%
  filter(continent == "Europe") %>%
  pull(correlation) %>% mean

```

We can see that most of the European countries truly achieve large negative correlations of
suicide rates and GDP. However, there are a few exceptions, where these two variables
are correlated positively.


Correlation between suicide rate and alcohol consumption.

```{r}
data_japan <- data %>% filter(country == "Japan")

data_japan %>% group_by(country, year) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = country)) +
  geom_line() +
  geom_point() +
  theme(legend.position = "none")

data_japan %>% group_by(country, year, age) %>%
  summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
  ) %>%
  ggplot(aes(x = year, y = suicides_per_100k, color = age)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = seq(1985, 2015, 2)) +
  labs(title = "Differences between age groups in Japan in time",
       x = "Year", y = "Suicides per 100k people")

```

### Conclusion




## Hypotheses Testing

Based on the previous analysis I would like to test several hypotheses.

### Suicide rates of men and women

The first hypothesis that I will test is that that men tend to commit
suicide more often than women. Therefore, I will test if the suicide rate of men
was higher than that of women during the time interval 1985-2015.

The null hypothesis is that
$$ \mu_m \leq \mu_w $$ 
and the alternative hypothesis is that
$$ \mu_m > \mu_w $$
where $\mu_m$ and $\mu_w$ are the means of suicide rates of men and women during
the time period 1985-2015, respectively. The confidence level is set to be $\alpha = 0.05$.

In order to test this hypothesis, I will make use of Welch's Two Sample t-test,
which is a modification of Student's t-test supporting unequal variances of the samples.

```{r}

sex.hypothesis.data <- data %>% group_by(year, sex) %>%
    summarise(
      population = sum(population),
      suicides_no = sum(suicides_no),
      suicides_per_100k = (suicides_no / population) * 100000,
      .groups = "keep"
    )

rates.male <- sex.hypothesis.data %>% filter(sex == "Male") %>% pull(suicides_per_100k)
rates.female <- sex.hypothesis.data %>% filter(sex == "Female") %>% pull(suicides_per_100k)

t.test(
  x = rates.male,
  y = rates.female,
  alternative = "greater"
)

```

We can see that the resulting p-value of the Welch's Two Sample t-test
is smaller than our confidence level, and therefore we reject the null hypothesis
in favor of the alternative hypothesis. We have confirmed our initial suspicion,
that is that men have higher suicide rates than women.


### Difference in suicide rates between age groups

Another interesting insight gained from the exploratory analysis was that there
is an apparent difference in suicide rates between individual age groups.





